---
output:
  pdf_document:
    includes:
      in_header: header.tex
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r}
title_var <- "MarineGEO Protocol"
```

---
title: `r title_var`
---

\noindent\rule{\textwidth}{0.4pt}

$$\includegraphics{title_pic}$$


\underline{How to cite this work:} Seagrass Habitats: MarineGEO Protocols. (2019) Tennenbaum Marine Observatories Network, MarineGEO, Smithsonian Institution. DOI:

\vspace*{\fill}
\begin{center}
\includegraphics[height=1.5cm, width=4.8cm]{smithsonian}\\
marinegeo@si.edu
\end{center}

\thispagestyle{empty}
\clearpage
\pagenumbering{arabic}

\noindent\rule{\textwidth}{0.4pt} 

## Introduction

\noindent\rule{\textwidth}{0.4pt}

## Measured Parameters

\noindent\rule{\textwidth}{0.4pt}

## Requirements

\underline{Survey Design:}

\renewcommand{\labelitemi}{$\square$}
\setlength{\parskip}{0.1em}
\begin{itemize}
  \item first item
  \item second item
\end{itemize}
\vspace{\baselineskip}

\underline{Fieldwork:}

\renewcommand{\labelitemi}{$\square$}
\setlength{\parskip}{0.1em}
\begin{itemize}
  \item first item
  \item second item
\end{itemize}
\vspace{\baselineskip}

\underline{Post-Processing:}

\renewcommand{\labelitemi}{$\square$}
\setlength{\parskip}{0.1em}
\begin{itemize}
  \item first item
  \item second item
\end{itemize}
\vspace{\baselineskip}

\noindent\rule{\textwidth}{0.4pt}

## Methods

Fully review this and any additional protocols necessary for the sampling excursion. Address any questions or concerns to marinegeo@si.edu before beginning this protocol.
\vspace{\baselineskip}

\underline{Preparation:}

\vspace{\baselineskip}
\underline{Fieldwork:}

\vspace{\baselineskip}
\underline{Post-Processing:}

\noindent\rule{\textwidth}{0.4pt}

## Data Submission

1. Scan the completed field data sheets and save both paper and electronic versions locally.
2. Enter data into provided data entry template.
3. Use our online submission portal to upload the Excel Spreadsheet (coming Fall 2019). 
4. Contact us if you have any questions: marinegeo@si.edu.

